---
title: "Income Growth in Urban vs Rural Counties"
author: "Macy Trout"
date: "September 2024"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    highlight: tango
    code_folding: hide
    df_print: paged
    css: styles.css
---

### Data Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r installling libraries, warning = FALSE, message = FALSE, echo = FALSE}
# tidycensus and load necessary libraries

library(tidycensus)
library(tidyverse)
library(ggplot2)
library(tigris)
library(dplyr)
library(sf)

# Census API key 
census_api_key("bddd8495974eca2afb75bb773c01c741730d44a5", install = TRUE, overwrite = TRUE)

```

```{r, warning = FALSE, message = FALSE, fig.show="hide", echo=FALSE}
# get 2010 ACS data
income_2010 <- get_acs(
  geography = "county",
  variables = "B19013_001", # Median household income variable
  year = 2010,
  survey = "acs5", 
  progress = FALSE
)

# get 2020 ACS data
income_2020 <- get_acs(
  geography = "county",
  variables = "B19013_001",
  year = 2020,
  survey = "acs5", 
  progress = FALSE
)

# rename variables
income_2010 <- income_2010 %>%
  rename(median_income_2010 = estimate)

income_2020 <- income_2020 %>%
  rename(median_income_2020 = estimate)

# merge 2010 and 2020 data by county (GEOID)
income_data <- inner_join(income_2010, income_2020, by = "GEOID")

# calculate percent change in income over the decade
income_data <- income_data %>%
  mutate(income_change = (median_income_2020 - median_income_2010) / median_income_2010 * 100)

# total population and land area for counties
population_data <- get_acs(
  geography = "county",
  variables = c(
    population = "B01003_001",
    land_area = "B01001_001"
  ),
  year = 2020,
  survey = "acs5", 
  progress = FALSE
)

# loading county shapefile with area
county_shapes <- counties(cb = TRUE)


# land area (make land area in square meters)
land_area_data <- county_shapes %>%
  select(GEOID, ALAND) %>%
  mutate(land_area_sq_miles = ALAND / 2.59e+6)

```


```{r, warning = FALSE, message = FALSE, fig.show="hide", echo=FALSE}
# merge population data with land area data using GEOID
population_with_land_area <- left_join(population_data, land_area_data, by = "GEOID")

# calculate population density
population_with_land_area <- population_with_land_area %>%
  mutate(population_density = estimate / land_area_sq_miles)

# counties based on population density
population_with_land_area <- population_with_land_area %>%
  mutate(classification = ifelse(population_density > 500, "Urban", "Rural"))

# merge the classification data with income data
income_data <- left_join(income_data, population_with_land_area, by = "GEOID")
```

```{r, warning = FALSE, message = FALSE, include = FALSE}
# county boundaries
county_shapes <- counties(cb = TRUE)

# merge income data with county shapefile by GEOID
county_income_map <- left_join(county_shapes, income_data, by = "GEOID")

# convert GEOID to character in both datasets
income_data <- income_data %>%
  mutate(GEOID = as.character(GEOID))

county_shapes <- county_shapes %>%
  mutate(GEOID = as.character(GEOID))

county_income_map <- left_join(county_shapes, income_data, by = "GEOID")

st_crs(county_shapes)

county_shapes <- st_transform(county_shapes, crs = 4326)

county_income_map <- county_income_map %>%
  filter(!STUSPS %in% c("AK", "HI"))

```
# A Growing Divide: How Urban and Rural America Drifted Further Apart in a Decade

Over the past decade, a silent yet profound transformation has swept across the United States. From 2010 to 2020, household incomes in counties across the country changed in ways that reveal a deepening divide—a divide that, if left unchecked, threatens to reshape the economic landscape of the nation for generations to come.

## A Nation Divided by Income Growth

To understand this shift, we first turn to the map of income changes. This map paints a striking picture of the growing divide between urban and rural counties in America.

*Urban America is booming.* The *deep red* hues in regions like the *West Coast, Northeast*, and parts of the *Southwest* reveal a surge in household incomes—rising by as much as *60%* in some counties. These are places where industries like technology, finance, and entertainment are flourishing, drawing talent and investment from around the world.

But this growth is not spread evenly. *Rural America tells a very different story.* Across much of the *Midwest* and *Southeast*, the map fades to *pale pink* and even *purple*—signifying income stagnation or, in some cases, outright decline. These counties, once the economic backbone of America’s agricultural and manufacturing industries, are being left behind.

```{r, warning = FALSE, message = FALSE}
# Refined ggplot map for income change with improved legend
ggplot(county_income_map) +
  # Add county geometries with borders and fill based on income change
  geom_sf(aes(fill = income_change, geometry = geometry.y), color = "white", size = 0.1) +
  
  # Use a diverging color palette with midpoint at 0 (neutral change)
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, 
                       name = "Income Change (%)", 
                       breaks = c(-20, 0, 20, 40, 60),  # Adjust to the data range
                       labels = c("-20%", "0%", "+20%", "+40%", "+60%")) +
  
  # Improve labels, titles, and captions
  labs(
    title = "Percent Change in Median Household Income (2010-2020)",
    subtitle = "Comparing Urban and Rural Counties",
    caption = "Source: ACS 2010 & 2020",
    fill = "Income Change (%)"
  ) +
  
  # Customize the theme with better font sizes, background, and legend position
  theme_minimal(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.title.position = "panel",
    plot.subtitle = element_text(face = "italic", size = 14, hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.95, face = "italic"),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    
    # Adjust legend position and text
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.key.height = unit(0.7, "cm"),  # Adjust the height of the legend keys
    legend.key.width = unit(0.3, "cm")    # Adjust the width of the legend keys
  ) +
  
  # Add guide for continuous color bar (optional tweak for better legend appearance)
  guides(fill = guide_colorbar(barwidth = 1.5, barheight = 7, title.position = "top", title.hjust = 0.5))

```

## Explaining the Divide: The Box Plot Tells All

The map gives us a broad overview, but the real story comes into focus when we dive deeper. Take a look at the box plot comparison of urban versus rural counties. Here, we can see the difference in income growth laid bare.

Urban Counties: The purple boxplot represents urban counties. Here, income growth is steady and predictable. Most urban counties have seen median household incomes rise by 10% to 20% over the past decade. There are fewer extreme outliers, meaning that while the gains in urban areas have been substantial, they have also been relatively stable. Urban counties have benefited from their proximity to economic hubs, infrastructure development, and the diversification of industries.

Rural Counties: The orange boxplot, on the other hand, tells a more volatile story. While some rural counties have indeed seen income growth comparable to urban areas, the overall picture is much more erratic. Several rural counties have seen income plummet by up to 50%, while others have enjoyed unprecedented growth, with outliers showing increases of over 100%. This wild variability underscores the precariousness of rural economies, which are often more reliant on a single industry or sector.

```{r, warning = FALSE, message = FALSE}

ggplot(income_data, aes(x = classification, y = income_change, fill = classification)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Urban" = "#D8BFD8", "Rural" = "#FFA07A")) +
  labs(title = "Distribution of Income Change: Urban vs Rural Counties (2010-2020)",
       x = "County Classification", y = "Percent Change in Median Income") +
  theme_minimal()

```

# A Deeper Look at the Regions

Both the map and the boxplot provide regional and statistical clarity, but they also raise important questions about why these shifts are happening.

*Urban Growth Hubs*: The *West Coast* is home to booming tech hubs like *San Francisco* and *Seattle*, while the *Northeast* boasts financial powerhouses like *New York City* and *Boston*. These areas have not only attracted young professionals and international talent, but they’ve also become magnets for investment and development. The result? A consistent upward trajectory in household incomes, as reflected in the tight, predictable distribution of the urban boxplot.

*Rural Struggles and Successes*: In contrast, rural counties, particularly in the *Midwest* and *Southeast*, show much more mixed results. The economies of these regions often rely heavily on industries like agriculture, mining, or manufacturing—sectors that have been hit hard by technological advances and globalization. While some rural areas, especially those near growing urban centers or tourist destinations, have benefited from spillover effects, many more are grappling with declining populations, aging infrastructure, and limited access to education and healthcare.


## Winners and Losers in the 10-Year Income Race

The gap between urban and rural counties is not just a quirk of geography—it’s the defining economic story of the last decade.

*Urban Winners*: Counties along the West Coast and in parts of the Northeast stand out as the *clear winners* of the past decade, enjoying income growth that outpaces much of the country. These areas are thriving, propelled forward by industries that are future-facing, adaptable, and resistant to economic downturns.

*Rural Losers*: Conversely, many rural counties have found themselves on the losing end of this shift. As global forces push toward urbanization and tech-based economies, rural areas have struggled to find their footing. The boxplot’s outliers—those rural counties with steep income declines—serve as a stark reminder that some regions are falling behind, unable to compete in the evolving economic landscape.

## The Consequences of a Divided Economy

What these visualizations make clear is that America is experiencing two very different economic realities: one of growth, prosperity, and opportunity in urban areas, and another of stagnation, uncertainty, and decline in rural communities.

The map and box plot together tell a story of a country moving in two directions. As urban counties continue to attract investment and talent, the divide between thriving cities and struggling rural regions will only grow. The *economic future of America*, it seems, may be defined not by what brings us together, but by what pulls us apart.




